---
title: "examples plots"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

Use of three datasets for data exploration:

*starwars
*mtcars
*storms

Unnesting the nested columns in starwars dataframe

```{r}
library(tidyr)

starwars %>% unnest(films) %>%
  unnest(starships)
  
```

Correlation of height and body weight 

```{r}
starwars %>%
  filter(mass < 1000) %>%
  ggplot(aes(x = height, y = mass, color = gender)) +
  geom_point() +
  geom_smooth(method = "lm")

starwas_nona <- starwars %>%
  drop_na()
  
cor(starwas_nona$height, starwas_nona$mass)

lm(starwas_nona$mass ~ starwas_nona$height)
```


Ordering the x axis with numerical months

```{r}
storms %>%
  ggplot(aes(x = as.character(month), y = pressure, color = status)) +
  geom_boxplot() +
  scale_x_discrete(limits = seq(1,9)) +
  xlab("Month")
```

Rotating the x axis labels

```{r}
storms %>%
  ggplot(aes(x = as.character(year), y = long)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Specifying aes() in the different functions/layers

```{r}
storms %>%
  ggplot(aes(x = lat, y = long)) + 
  geom_point(aes(color = status)) +
  geom_smooth()

storms %>%
  ggplot(aes(x = lat, y = long)) + 
  geom_point(color = 'red') +
  geom_smooth()
```

Converting a numeric coloring to a cathegorical coloring

```{r}
mtcars %>%
  ggplot(aes( x = disp, y = wt, color = as.factor(cyl))) +
  geom_point()
```

Difference between & and && operators

```{r}
a <- c(TRUE,TRUE,TRUE)
b <- c(FALSE,FALSE,TRUE)

a & b
a && b

a | b
a || b
```
