---
title: "Introduction data handling and graphical representation"
author: Katharina Imkeller
date: 15. June 2020
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DT)
```

# Data wrangling with dplyr

dplyr is an R package that allows easy manipulation of data tables.
A set of commonly used dplyr commands can be found on the following 
summary: [dplyr cheatsheet](./data-transformation_cheatsheet.pdf).

The dplyr functions are:

* filter() picks cases based on their values.
* summarise() reduces multiple values down to a single summary.
* arrange() changes the ordering of the rows.
* mutate() adds new variables that are functions of existing variables
* select() picks variables based on their names.

The following script illustrate how to use these functions on a dataset 
containing information on different types of iris plants. This dataset 
is just for illustration purposes, because it is especially well suited 
to understand how dplyr is working. A similar introduction using a dataset 
about Starwars characters can also be found here: https://dplyr.tidyverse.org/.

Later on in the tutorial you will perform the same type of manipulations 
on a dataset of antibody reactivity data.

```{r}
library(dplyr)
```


## Filtering data tables

filter() picks cases (rows) based on their values.
In all following example, the DT::datatable() function is there to display the
tables in a readable way.

```{r}
DT::datatable(
filter(iris, Species == "versicolor")
)
```

Usage of the %>% pipe:
```{r}
DT::datatable(
iris %>%
    filter(Species == "versicolor")
)
```

Filters can be combined:
```{r}
DT::datatable(
iris %>%
    # AND
    filter(Species == "versicolor" & Petal.Width < 1.2) 
)
```

```{r}
DT::datatable(
iris %>%
    #  OR
    filter(Species == "versicolor" | Petal.Width < 1.2) 
)
```

# Summarizing and grouping

summarise() reduces multiple values down to a single summary

```{r}
DT::datatable(
iris %>%
  summarise(
    n = n(),
    meanSW = mean(Sepal.Width, na.rm = TRUE)
  ) 
)
```

```{r}
DT::datatable(
iris %>%
  group_by(Species) %>%
    summarise(
    n = n(),
    meanSW = mean(Sepal.Width, na.rm = TRUE)
  )
)
```


arrange() changes the ordering of the rows

```{r}
DT::datatable(
iris %>%
  group_by(Species) %>%
    summarise(
    n = n(),
    meanSW = mean(Sepal.Width, na.rm = TRUE)
  ) %>%
    arrange(meanSW)
)
```

Obtain the reverse order by adding the desc() function

```{r}
DT::datatable(
iris %>%
  group_by(Species) %>%
    summarise(
    n = n(),
    meanSW = mean(Sepal.Width, na.rm = TRUE)
  ) %>%
    arrange(desc(meanSW))
)
```

## Selecting columns

select() picks variables based on their names.

```{r}
DT::datatable(
iris %>%
  select(Sepal.Width, Species)
)
```

```{r}
DT::datatable(
iris %>%
  select(Species, ends_with("Width"))
)
```


## Adding columns (mutating the table)

mutate() adds new variables that are functions of existing variables

```{r}
DT::datatable(
iris %>%
  mutate(Sepal.Ratio = Sepal.Width/Sepal.Length) 
)
```


# Generating graphics using ggplot2

ggplot2 is an R package for generating graphical representations of your data.
A set of commonly used dplyr commands can be found on the following 
summary: [ggplot2 cheatsheet](./data-visualization-2.1_cheatsheet.pdf).

```{r}
library(ggplot2)
```

A ggplot2 command consists of following elements:

* a ggplot() function that has the dataset  (here iris) as first argument
* an aesthetic mapping aes(), where you define the columns of the dataset 
that should be used as x, y, color, size...
* a layer, where you define what type of plotting functions should be used 
(layers can be overlayed)

The following code plots the sepal length as a function of sepal length. 
Every dot corresponds to one row of the dataset iris.

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
    geom_point()
```

The dots can be colored according to iris species:

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color=Species)) +
    geom_point()
```

A faceting specification (e.g. facet_wrap()) can be used to split the data into several plots:

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color=Species)) +
    geom_point() +
    facet_wrap(~Species)
```

The data can also be summarized into box-plots, by using the Species on the x axis:

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length)) +
    geom_boxplot()
```

Overlaying dot-plots on top of the box-plot:

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length)) +
    geom_boxplot() +
    geom_jitter()
```



