---
title: "Automated analysis of antibody reactivity data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dataset on antibodies in human Malaria

The file csp_antibodies.csv contains sequence information and ELISA data from 178 antibodies from donors undergoing controlled human malaria infection [Murugan et al. 2010](https://www.nature.com/articles/s41591-020-0881-9). The Circumsporozoite Protein (CSP) is the major surface antigen of Plasmodium falciparum (Pf), the most frequent causative pathogen of human Malaria.

### General questions concerning the dataset

With the help of the original publication by [Murugan et al. 2010](https://www.nature.com/articles/s41591-020-0881-9), and with the knowledge from the lecture on B cells, please answer the following questions.

CHMI and ELISA:

* What is controlled human malaria infection (CHMI)? What do the different entries in the "Pf_exposure" column stand for (I, II, III, C)?
* Shortly summarize, how the ELISA was performed. What does the are under the curve (AUC) stand for?
* List the different CSP-derived epitopes that are used in the ELISA measurements. What are their full protein sequences?
* What information does the column Pf_inhibition contain, and how was it measured?

Antibody sequence information:

* What is the overall structure of an antibody molecule? How is the protein encoded in the B cell genome?
* Which biological processes underly the diversity of antibody molecules?
* What do the following columns in the dataset stand for: IGHV, IGHD, IGHJ, light, IGKV_IGLV, IGKJ_IGLJ, IGHC, IGHV_SHM, IGKV_IGLV_SHM? (Hint: column "light" indicated, whether the antibody has a kappa or lambda light chain, light chains can be encoded in IGK or IGL loci.)


### Exploratory data analysis

Load the libraries for dplyr and ggplot2:

```{r}

```

Read the csv table into a table object (use the read.csv() function):

```{r}

```

### Antibody sequence information, repertoire analysis

How many antibodies are there that have a kappa light chain? How many have a lambda light chain?

```{r}

```

How many antibodies does the dataset contain per donor?

```{r}

```

Which is the most prominent IGHV segment?

```{r}

```

What is the most frequent combination of IGHV and IGKV or IGLV genes?

```{r}

```

Antibody isotype, or constant segment (IGHC): using the summary function followed 
by a ggplot2 command with the geom_col() function, 
plot the number of antibodies from each isotype.
Which is the most frequent IGHC segment in the dataset?

```{r}

```

### ELISA reactivity to different epitopes

For the different antigens tested in ELISA, which donors show the most reactive antibodies?
Use geom_boxplot() and/or geom_jitter().

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

Using the filter() function and an AUC threshold of 5, 
can you extract the two groups of antibodies shown in Figure 1c, pannels 1 and 2 
(NPDN/NVDP binders and NVDP/NANP binders)?

How do these two groups differ in terms of their light chain usage and IGHV usage?

```{r}

```

```{r}

```

```{r}

```



### Pf inhibition

Plot the Pf inhibition for every donor (geom_boxplot() or geom_jitter() functions).
Which donor has the most protective antibodies?

```{r}

```

Using the cor() function within a summarize command, calculate the
correlation of Pf inhibition and ELISA AUC for every antigen.

```{r}

```

Plot the correlation of Pf inhibition and NANP for the
highest and lowest correlation found in the previous calculation.

```{r}

```

```{r}

```

## Summary and conclusion

Shortly describe the typical sequence properties of anti-CSP antibodies that you found in part 1 of the
above analysis.

Considering the second part of the analysis, which donors seem to have the best antibody
response? Which of the CSP epitopes seems to bound most frequently by protective antibodies?
