---
title: "Introduction to R and Bioconductor"
author: Katharina Imkeller
date: 15. June 2020
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this course we will learn how to analyse large scale immunological data using R. 
R is a programming language and environment for scientific computing and graphics generation. 
A major aim of this course is to enable students to perform basic analyses on their own 
biological datasets. To this purpose will introduce various R community resources that 
will allow students to get support and further develop their programming skills.

# RStudio and basic R commands

## Connecting to RStudio cloud

In this course we will program our R scripts in an RStudio cloud environment. 
All packages that are needed for the course have been installed there.
Please follow the link to the Course Space: 

https://rstudio.cloud/spaces/73774/join?access_code=OicSZ6BeiynxBC7X%2F6VetC4OWstPLt9Hx330n08c

You will have to register there, by creating an account or using an account of your choice, e.g. a Google account.

## Basic R commands

Simple calculations
```{r}
4 + 6
```

Assigning variables
```{r}
a <- 4
b <- 6
a + b
```

Functions use variables to perform a computation and return the result
```{r}
myfunction <- function(x1, x2) {
    x1 + x2
}

myfunction(4, 6)

myfunction(a, b)
```

Predefined functions
```{r}
seq(1, 10)
```

Getting help for a specific function
```{r}
?seq
```


# Rmarkdown

[Rmarkdown](https://rmarkdown.rstudio.com/) is a notebook interface, which allows 
the generation of reproducible and sharable report of your R analyses. The format of 
an Rmarkdown allows to integrate formatted text, computational code, calculations and 
results as well as graphical representations into the same document. The output 
document can be generated as pdf or html file for example. In fact most of the materials 
used in this course use Rmarkdown format. 

Formatting guidelines for Rmarkdown can be found in this [Rmarkdown cheatsheet](./rmarkdown-2.0_cheatsheet.pdf).

# Bioconductor

Bioconductor is an open source software collection for Bioinformatics. 
It contains more than 1800 R packages for statistical analysis and comprehension 
of high-throughput biological data (transcriptomics, genomics, flow cytometry, 
proteomics, metabolomics...). Bioconductor packages are widely used 
(>1/2 million unique IP downloads / year) and highly cited (>33,000 PubMedCentral citations). 
In order to submit packages to the Bioconductor collection, they have to go through a 
review process that assures quality control.

The Bioconductor community offers a large number of open source resources to learn using 
packages and analyzing data. Typical biological data analysis workflows such as RNAseq or 
proteomics for example can be found on the [workflow website](https://bioconductor.org/packages/release/BiocViews.html#___Workflow).

## Using the Bioconductor Biostrings package

[Biostrings](https://bioconductor.org/packages/release/bioc/html/Biostrings.html) is a package that 
allows you to perform basic operations on biological sequences

```{r}
dnasequence <- "ATGCAGTACGATCAGCAGTCAACCGTGGGGGGGACCCAG"
```

Loading the Biostrings package makes the package functions available in your programming environment.

```{r, message=FALSE}
library(Biostrings)
```

Define the above sequence as a DNA sequence

```{r}
dnaobj <- DNAString(dnasequence)
```

Translate the DNA sequence into protein sequence

```{r}
translate(dnaobj)
```

# R tutorials for further reading

Beginner: https://cran.r-project.org/doc/contrib/Torfs+Brauer-Short-R-Intro.pdf

Advanced: https://r4ds.had.co.nz/workflow-projects.html